Dissecting the relationship between protein structure and sequence variation

### 2015-06-05 manuscript/voronoi/mnv.tex
MN update

    In order to segregate the combined effects of long-range interactions from local packing density, the inverse of the Voronoi cell volume can be used as a maximally local measure of packing density. By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the effects long-range amino acid interactions beyond immediate neighbors -- on sequence variability. Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume. It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman $\rho$ is a non-parametric rank correlation coefficient.

    The resulting distribution of the Spearman's partial correlation coefficients of evolutionary rates with wcn (calculated using side-chain coordinates, wcnSC) has an absolute median value of 0.32 with a $50\%$ quartile range of $[0.23,0.40]$ about the median of the distribution.

@@

    Throughout the previous sections,  a comprehensive analysis and comparison of the main structural determinants of sequence variability was carried out, using a dataset of $209$ monomeric enzymes. Examples of sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number, Relative Solvent Accessibility (RSA), and \ddg rate as defined in Chapter \ref{chap:sschar} (see also Echave et al. (2014) \cite{echave_relationship_2014}), which is essentially a proxy measure of the stability of protein's native conformation upon substitution of amino acids in individual sites in proteins. In addition, we have derived new site-specific characteristics from the Voronoi Tessellation of protein 3D structures, that are capable of explaining sequence variability equally well or better than several previously considered structural quantities, such as B factor, RSA, \ddg rate, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (e.g., Figures \ref{fig:best_predictorER} \& \ref{fig:best_predictorSE}).

>>>>

    In order to segregate the combined effects of long-range interactions from local packing density, the inverse of the Voronoi cell volume can be used as a maximally local measure of packing density. By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the effects long-range amino acid interactions beyond immediate neighbors -- on sequence variability. Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume. It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman $\rho$ is a non-parametric rank correlation coefficient. \\

    The resulting distribution of the Spearman's partial correlation coefficients of evolutionary rates with wcn (calculated using side-chain coordinates, wcnSC) has an absolute median value of $0.32$ with a $50\%$ quartile range of $[0.23,0.40]$ about the median of the distribution. Therefore, the long-range amino acid interactions appear to explain approximately $10\%$ of the site-specific sequence evolutionary rates. The local packing density as measured by the inverse of Voronoi cell volume is alone capable of explaining a median $35\%$ of sequence evolutionary rates, corresponding to a median Spearman's correlation strength of $0.59$ with $50\%$ quartile range of $[0.52,0.64]$. By contrast, the partial correlation distribution of RSA while controlling for the Voronoi cell volume for the same dataset, indicates a negligible median contribution of $\sim0.006$ in explaining the observed variability in evolutionary rates of individual proteins.

@@

    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most important structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes. Examples of important sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA). In addition, we have derived a new set of site-specific characteristics from the Voronoi Tessellation of protein 3D structures, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).

    We also argued that the commonly used definitions of local packing density as measured by the Contact Number (CN) and the Weighted Contact Number (WCN), involve a free adjustable parameter in the definition. In order to get the best predictions of sequence variability, we fine-tuned the free parameters of the models for each proteins individually, and showed that the best performing parameter values of CN and WCN indicate a non-negligible influence of the long-range amino acid interactions, independently of the local packing density.
